---
layout: page
title: Intermediate programming with R
subtitle: Debugging with browser
minutes: 30
---

```{r, include = FALSE}
source("tools/chunk-options.R")
```

> ## Learning Objectives {.objectives}
>
> *  Use `browser` to set breakpoints
> *  Use `browser` to set conditional breakpoints

In the last lesson we used `debug` to enter into a function's environment for interactive debugging.
However, if we have an idea where the bug is located, we can use the function `browser` to set a "breakpoint" in that location.
This prevents us from having to step through each line of a function to reach the point where the problem is located.
Furthermore, we can use a conditional statement to only activate the debugger when a certain condition is true (especially useful for long `for` loops).

> ## Turn off RStudio's debugging features {.callout}
>
> Just like last lesson, make sure to turn of RStudio's debugging features.
> In the menu, go to "Debug".
> From the dropdown menu, go to "On Error" and choose the setting "Message Only".


```{r echo = FALSE}
counts_raw <- read.delim("data/counts-raw.txt.gz")
```

We'll start with the function we updated in the last lesson.

```{r}
mean_metric_per_var <- function(metric, variable) {
  if (!is.factor(variable)) {
    variable <- as.factor(variable)
  }
  variable <- droplevels(variable)
  result <- numeric(length = length(levels(variable)))
  names(result) <- levels(variable)
  for (v in levels(variable)) {
    result[v] <- mean(metric[variable == v])
  }
  return(result)
}
```

And we'll focus on fixing the following behavior:

```{r}
mean_metric_per_var(counts_raw$facebookLikeCount, counts_raw$journal)
```

Our function returns `NA` for the total number of Facebook likes for PLOS One.
Why is this happening?
It is correctly identifying all the journals, so the first section of code appears to be working correctly.
The totals per journal are computed within the `for` loop, so that is likely where the problem is originating.
Since we suspect the problem is occuring during the `for` loop, we'll set the breakpoint there with `browser` instead of starting from the beginning of the function using `debug`.

```{r}
mean_metric_per_var <- function(metric, variable) {
  if (!is.factor(variable)) {
    variable <- as.factor(variable)
  }
  variable <- droplevels(variable)
  result <- numeric(length = length(levels(variable)))
  names(result) <- levels(variable)
  for (v in levels(variable)) {
    browser()
    result[v] <- mean(metric[variable == v])
  }
  return(result)
}
```

Now the next time we call the function, we are dropped into the debugger at the breakpoint set by `browser`.

```{r eval=FALSE}
mean_metric_per_var(counts_raw$facebookLikeCount, counts_raw$journal)
```
~~~ {.output}
Called from: mean_metric_per_var(counts_raw$facebookLikeCount, counts_raw$journal)
~~~
~~~ {.r}
Browse[1]> 
~~~

Let's confirm that the beginning of the function has already been run.

~~~ {.r}
Browse[1]> ls()
~~~
~~~ {.output}
[1] "metric"   "result"   "v"        "variable"
~~~

Furthermore, we can check the current value of `v` in the `for` loop.

~~~ {.r}
Browse[1]> v
~~~
~~~ {.output}
[1] "pbio"
~~~

> ## Investigating variables with the same names as debugging commands {.callout}
> What if you use one of the debugging commands, e.g. `n`, `s`, `f`, as one of the names of your variables?
> If you type them into the debugger, they will be used as commands.
> If you want to view the value contained by the variable in the debugger, use `print`, e.g. `print(n)`.

We want to see what is happening when `v` is `"pone"`.
As we did before we could step through line by line using `n`.
But using this approach, each time through the loop we would have to type `n` multiple times to run the lines of code in the loop.
This would be even worse if there were many lines of code inside.
Instead, we can use `c` for "continue", which continues running the code until the next time `browser` is called.

~~~ {.r}
Browse[1]> c
~~~
~~~ {.output}
Called from: mean_metric_per_var(counts_raw$facebookLikeCount, counts_raw$journal)
~~~
~~~ {.r}
Browse[1]> v
~~~
~~~ {.output}
[1] "pcbi"
~~~
~~~ {.r}
Browse[1]> c
~~~
~~~ {.output}
Called from: mean_metric_per_var(counts_raw$facebookLikeCount, counts_raw$journal)
~~~
~~~ {.r}
Browse[2]> v
~~~
~~~ {.output}
[1] "pgen"
~~~

But this really isn't much better, especially if we run through the `for` loop multiple times as we attempt to debug the function.
Let's quit the debugger and try a new strategy.

~~~ {.r}
Browse[2]> Q
~~~

We can set a conditional breakpoint using an `if` statement.
Then we will be dropped into the interactive debugger only when the condition is true.
We want to enter the debugger when `v == "pone"`.

```{r}
mean_metric_per_var <- function(metric, variable) {
  if (!is.factor(variable)) {
    variable <- as.factor(variable)
  }
  variable <- droplevels(variable)
  result <- numeric(length = length(levels(variable)))
  names(result) <- levels(variable)
  for (v in levels(variable)) {
    if (v == "pone") {
      browser()
    }
    result[v] <- mean(metric[variable == v])
  }
  return(result)
}
```

```{r eval=FALSE}
mean_metric_per_var(counts_raw$facebookLikeCount, counts_raw$journal)
```
~~~ {.output}
Called from: mean_metric_per_var(counts_raw$facebookLikeCount, counts_raw$journal)
~~~
~~~ {.r}
Browse[1]> v
~~~
~~~ {.output}
[1] "pone"
~~~

Now we entered into the debugger after the loop has reached "pone".
Let's inspect the variable the values being passed to `mean`.
Specifically, let's see all the unique values.

~~~ {.r}
Browse[1]> unique(metric[variable == v])
~~~
~~~ {.output}
 [1]   0   2   1   6   5   3  10   7   4   8  35  12  34   9  37  16  14  25 892  22
[21]  11  19  18  13  21  49  15  41  50  17  51 104  23  20  26 151  30  39  NA  24
[41]  95  44 109  66
~~~

Interestingly, at least one of the Facebook Like counts are NA.
Is this different from the other journals?
Let's check "pbio".

~~~ {.r}
Browse[1]> unique(metric[variable == "pbio"])
~~~
~~~ {.output}
 [1]  0  1  2  3  4 12  8 39  7  5 10 45 11 18 44  6 14 13 43 19
~~~

It does not contain any `NA`s, so this is likely the problem.

Let's first exit the debugging environment.

~~~ {.r}
Browse[1]> Q
~~~

And then check the help for `mean` to see if we can figure out what is going on (remember you can also press the `F1` key to see a function's help page).

```{r eval=FALSE}
?mean
```

From the help page, we see that `mean` has an argument `na.rm` to remove `NA`s.

> na.rm	a logical value indicating whether NA values should be stripped before the computation proceeds.

Otherwise by default, the function returns `NA` if any of the values are `NA`.

Let's update the function so that `mean` removes `NA` values.
At this point we can also remove the call to `browser`.

```{r}
mean_metric_per_var <- function(metric, variable) {
  if (!is.factor(variable)) {
    variable <- as.factor(variable)
  }
  variable <- droplevels(variable)
  result <- numeric(length = length(levels(variable)))
  names(result) <- levels(variable)
  for (v in levels(variable)) {
    result[v] <- mean(metric[variable == v], na.rm = TRUE)
  }
  return(result)
}
```

And now the function works properly when passed an `NA`!

```{r}
mean_metric_per_var(counts_raw$facebookLikeCount, counts_raw$journal)
```
