---
layout: page
title: Intermediate programming with R
subtitle: Debugging with browser
minutes: 30
---

```{r, include = FALSE}
source("tools/chunk-options.R")
```

> ## Learning Objectives {.objectives}
>
> *  Use `browser` to set breakpoints
> *  Use `browser` to set conditional breakpoints

In the last lesson we used `debug` to enter into a function's environment for interactive debugging.
However, if we have an idea where the bug is located, we can use the function `browser` to set a "breakpoint" in that location.
This prevents us from having to step through each line of a function to reach the point where the problem is located.
Furthermore, we can use a conditional statement to only activate the debugger when a certain condition is true (especially useful for long `for` loops).

> ## Turn off RStudio's debugging features {.callout}
>
> Just like last lesson, make sure to turn of RStudio's debugging features.
> In the menu, go to "Debug".
> From the dropdown menu, go to "On Error" and choose the setting "Message Only".


```{r echo = FALSE}
counts_raw <- read.delim("data/counts-raw.txt.gz")
```

We'll start with the function we updated in the last lesson.

```{r}
sum_metric_per_var <- function(metric, variable) {
  if (!is.factor(variable)) {
    variable <- as.factor(variable)
  }
  variable <- droplevels(variable)
  result <- numeric(length = length(levels(variable)))
  names(result) <- levels(variable)
  for (v in levels(variable)) {
    result[v] <- sum(metric[variable == v])
  }
  return(result)
}
```

And we'll focus on fixing the following behavior:

```{r}
sum_metric_per_var(counts_raw$facebookLikeCount, counts_raw$journal)
```

Our function returns `NA` for the total number of Facebook likes for PLOS One.
Why is this happening?
Since we know that the problem is occuring during the `for` loop, we'll set the breakpoint there with `browser` instead of starting from the beginning of the function using `debug`.

```{r}
sum_metric_per_var <- function(metric, variable) {
  if (!is.factor(variable)) {
    variable <- as.factor(variable)
  }
  variable <- droplevels(variable)
  result <- numeric(length = length(levels(variable)))
  names(result) <- levels(variable)
  for (v in levels(variable)) {
    browser()
    result[v] <- sum(metric[variable == v])
  }
  return(result)
}
```

Now the next time we call the function, we are dropped into the debugger at the breakpoint set by `browser`.

```{r eval=FALSE}
sum_metric_per_var(counts_raw$facebookLikeCount, counts_raw$journal)
```
~~~ {.output}
Called from: sum_metric_per_var(counts_raw$facebookLikeCount, counts_raw$journal)
~~~
~~~ {.r}
Browse[1]> 
~~~

Let's confirm that the beginning of the function has already been run.

~~~ {.r}
Browse[1]> ls()
~~~
~~~ {.output}
[1] "metric"   "result"   "v"        "variable"
~~~

Furthermore, we can check the current value of `v` in the `for` loop.

~~~ {.r}
Browse[1]> v
~~~
~~~ {.output}
[1] "pbio"
~~~

> ## Investigating variables with the same names as debugging commands {.callout}
> What if you use one of the debugging commands, e.g. `n`, `s`, `f`, as one of the names of your variables?
> If you type them into the debugger, they will be used as commands.
> If you want to view the value contained by the variable in the debugger, use `print`, e.g. `print(n)`.

We want to see what is happening when `v` is `"pone"`.
As we did before we could step through line by line using `n`.
But using this approach, each time through the loop we would have to type `n` multiple times to run the lines of code in the loop.
This would be even worse if there were many lines of code inside.
Instead, we can use `c` for "continue", which continues running the code until the next time `browser` is called.

~~~ {.r}
Browse[1]> c
~~~
~~~ {.output}
Called from: sum_metric_per_var(counts_raw$facebookLikeCount, counts_raw$journal)
~~~
~~~ {.r}
Browse[1]> v
~~~
~~~ {.output}
[1] "pcbi"
~~~
~~~ {.r}
Browse[1]> c
~~~
~~~ {.output}
Called from: sum_metric_per_var(counts_raw$facebookLikeCount, counts_raw$journal)
~~~
~~~ {.r}
Browse[2]> v
~~~
~~~ {.output}
[1] "pgen"
~~~

But this really isn't much better, especially if we run through the `for` loop multiple times as we attempt to debug the function.
Let's quit the debugger and try a new strategy.

~~~ {.r}
Browse[2]> Q
~~~

We can set a conditional breakpoint using an `if` statement.
Then we will be dropped into the interactive debugger only when the condition is true.
We want to enter the debugger when `v == "pone"`.

```{r}
sum_metric_per_var <- function(metric, variable) {
  if (!is.factor(variable)) {
    variable <- as.factor(variable)
  }
  variable <- droplevels(variable)
  result <- numeric(length = length(levels(variable)))
  names(result) <- levels(variable)
  for (v in levels(variable)) {
    if (v == "pone") browser()
    result[v] <- sum(metric[variable == v])
  }
  return(result)
}
```

```{r eval=FALSE}
sum_metric_per_var(counts_raw$facebookLikeCount, counts_raw$journal)
```
~~~ {.output}
Called from: sum_metric_per_var(counts_raw$facebookLikeCount, counts_raw$journal)
~~~
~~~ {.r}
Browse[1]> v
~~~
~~~ {.output}
[1] "pone"
~~~

Now we entered into the debugger after the loop has reached "pone".
Let's inspect the variable the values being passed to `sum`.
Specifically, let's see all the unique values.

~~~ {.r}
Browse[1]> unique(metric[variable == v])
~~~
~~~ {.output}
 [1]   0   2   1   6   5   3  10   7   4   8  35  12  34   9  37  16  14  25 892  22
[21]  11  19  18  13  21  49  15  41  50  17  51 104  23  20  26 151  30  39  NA  24
[41]  95  44 109  66
~~~

Interestingly, at least one of the Facebook Like counts are NA.
Is this different from the other journals?
Let's check "pbio".

~~~ {.r}
Browse[1]> unique(metric[variable == "pbio"])
~~~
~~~ {.output}
 [1]  0  1  2  3  4 12  8 39  7  5 10 45 11 18 44  6 14 13 43 19
~~~

It does not contain any `NA`s.
So let's check the help for `sum` to see if we can figure out what is going on (remember you can also press the `F1` key to see a function's help page).

```{r eval=FALSE}
?sum
```

From the help page, we see that `sum` has an argument `na.rm` to remove `NA`s.

> na.rm	logical. Should missing values (including NaN) be removed?

Furthermore, the function returns `NA` of any of the values are `NA`.

> If na.rm is FALSE an NA or NaN value in any of the arguments will cause a value of NA or NaN to be returned, otherwise NA and NaN values are ignored.

Let's update the function so that `sum` removes `NA` values before summing the vector.
At this point we can also remove the call to `browser`.

```{r}
sum_metric_per_var <- function(metric, variable) {
  if (!is.factor(variable)) {
    variable <- as.factor(variable)
  }
  variable <- droplevels(variable)
  result <- numeric(length = length(levels(variable)))
  names(result) <- levels(variable)
  for (v in levels(variable)) {
    result[v] <- sum(metric[variable == v], na.rm = TRUE)
  }
  return(result)
}
```

And now the function works properly when passed an `NA`!

```{r}
sum_metric_per_var(counts_raw$facebookLikeCount, counts_raw$journal)
```
