---
layout: page
title: Intermediate programming with R
subtitle: Creating different plots with geoms
minutes: 30
---

```{r, include = FALSE}
source("tools/chunk-options.R")
opts_chunk$set(fig.path = "fig/00-ggplot2-geoms-")
```

> ## Learning Objectives {.objectives}
>
> * Map a categorical (discrete) variable to an axis
> * Create a boxplot and barplot
> * Add error bars and text annotations

To this point we have mainly focused on a scatter plot created using `geom_point`.
ggplot2 offers many different geoms to create a wide variety of plots.
In this lesson, we'll explore a few more geoms.

```{r echo=FALSE}
counts_raw <- read.delim("data/counts-raw.txt.gz")
```

```{r echo=FALSE, message=FALSE}
library("dplyr")
```

```{r echo=FALSE}
research <- filter(counts_raw, articleType == "Research Article")
```

```{r echo=FALSE, message=FALSE}
library("ggplot2")
```

### Mapping categorical (discrete) variables

The scatter plot we made mapped continuous variables to the x and y axes.

```{r}
p <- ggplot(research, aes(x = log10(pdfDownloadsCount + 1), y = log10(wosCountThru2011 + 1))) +
  geom_point(aes(color = journal)) +
  geom_smooth() +
  scale_x_continuous(breaks = c(1, 3), labels = c(10, 1000)) +
  scale_y_continuous(breaks = c(1, 3), labels = c(10, 1000))
p
```

Of course we can also map categorical (also known as discrete) variables.

```{r}
p_box <- ggplot(research, aes(x = journal, y = log10(wosCountThru2011 + 1))) +
  geom_boxplot() +
  scale_y_continuous(breaks = c(1, 3), labels = c(10, 1000))
p_box
```

### Creating a barplot

A common figure in scientific publications is the barplot, where the height of the bars represent the mean and the error bars represent the standard error of the mean (SEM).
Recall from the challenge in the dplyr section that we calculated the mean and SEM for the number of article tweets per journal.

```{r}
tweets_per_journal <- research %>%
  group_by(journal) %>%
  summarize(num = n(),
            mean = mean(backtweetsCount),
            sem = sd(backtweetsCount) / sqrt(num))
tweets_per_journal
```

Let's create a barplot to display this result using `geom_bar`.

```{r}
tweets_bar <- ggplot(tweets_per_journal, aes(x = journal, y = mean)) +
  geom_bar(stat = "identity")
tweets_bar
```

We have to set the argument `stat = "identity"` because by default `geom_bar` tabulates the number of occurrences of the variable mapped to `x` and maps this frequency to `y`.

Now let's add the error bars using `geom_errorbar`.
We pass it values for `ymin` and `ymax`, which define the range of the error bars.

```{r}
tweets_bar <- ggplot(tweets_per_journal, aes(x = journal, y = mean)) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin = mean - sem, ymax = mean + sem))
tweets_bar
```

By default the error bars are as wide as the bars.
Let's reduce their size using the argument `width`.

```{r}
tweets_bar <- ggplot(tweets_per_journal, aes(x = journal, y = mean)) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin = mean - sem, ymax = mean + sem), width = 0.1)
tweets_bar
```

Since we also calculated the sample size per journal, let's add the number of articles to the plot using `geom_text`.

```{r}
tweets_bar <- ggplot(tweets_per_journal, aes(x = journal, y = mean)) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin = mean - sem, ymax = mean + sem), width = 0.1) +
  geom_text(aes(label = num))
tweets_bar
```

These are difficult to read because the text boxes are centered on the coordinate that marks the top of each bar.
The positioning of the text box is controlled by the arguments `hjust` and `vjust`, for horizontal and vertical justification.
By default these are both set to 0.5.
We can image the text box being defined from 0 to 1 from left to right for `hjust` and from 0 to 1 from bottom to top for `vjust`.
The values we set for `hjust` and `vjust` adjust where the text box is situated relative to that coordinate at the top of each bar.
We'll set them both to 0 to place the bottom left corner of the text box at that point, and thus the text is above and to the right of the bar.

```{r}
tweets_bar <- ggplot(tweets_per_journal, aes(x = journal, y = mean)) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin = mean - sem, ymax = mean + sem), width = 0.1) +
  geom_text(aes(label = num), hjust = 0, vjust = 0)
tweets_bar
```

To make it easier to see this central point, and to show an alternative to a boxplot, we can easily switch from a barplot to using points.

```{r}
tweets_point <- ggplot(tweets_per_journal, aes(x = journal, y = mean)) +
  geom_point() + 
  geom_errorbar(aes(ymin = mean - sem, ymax = mean + sem), width = 0.1) +
  geom_text(aes(label = num), hjust = 0, vjust = 0)
tweets_point
```

Using multiple geoms, we were able to created a plot with multiple layers to display multiple sources of information, the mean, SEM, and sample size.

### Challenges

> ## Mean number of tweets per journal per year {.challenge}
>
> Modify the dplyr code above to calculate the mean, SEM, and sample size of the number of article tweets per journal _and_ per year.
> Use `facet_wrap` to make a separate subplot per year.

```{r include=FALSE}
tweets_per_journal_year <- research %>%
  group_by(journal, year) %>%
  summarize(num = n(),
            mean = mean(backtweetsCount),
            sem = sd(backtweetsCount) / sqrt(num))
tweets_per_journal_year

ggplot(tweets_per_journal_year, aes(x = journal, y = mean)) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin = mean - sem, ymax = mean + sem), width = 0.1) +
  geom_text(aes(label = num), hjust = 0, vjust = 0) +
  facet_wrap(~year)
```

> ## Visualizing a single distribution {.challenge}
>
> The geoms `geom_histogram` and `geom_density` can be used to create histograms and density plots, respectively.
> Using these geoms, visualize the distribution of 2011 citations (`wosCountThru2011`).
> Compare the raw distribution to log10 and square root transformations.

```{r include=FALSE}
p <- ggplot(research, aes(x = wosCountThru2011))
p + geom_histogram()
p + geom_histogram() + scale_x_log10() 
p + geom_histogram() + scale_x_sqrt()
p + geom_density()
p + geom_density() + scale_x_log10() 
p + geom_density() + scale_x_sqrt()
```
