---
layout: page
title: Intermediate programming with R
subtitle: ggplot2
minutes: 10
---

> ## Learning Objectives {.objectives}
>


```{r}
library("dplyr")
library("ggplot2")
```

```{r}
counts_norm <- read.delim("data/counts-norm.txt.gz", stringsAsFactors = FALSE)
```


Do papers with more authors get more citations?

```{r}
ggplot(counts_norm, aes(x = authorsCount, y = wosCountThru2011)) +
  geom_point()
```

Do papers with more authors get more social media buzz?

```{r}
ggplot(counts_norm, aes(x = authorsCount, y = backtweetsCount)) +
  geom_point()
```

```{r}
ggplot(counts_norm, aes(x = authorsCount, y = facebookLikeCount)) +
  geom_point()
```

Relationship between Twitter and Facebook.

```{r}
ggplot(counts_norm, aes(x = backtweetsCount, y = facebookLikeCount)) +
  geom_point()
```

Number of authors by journal.

```{r}
ggplot(counts_norm, aes(x = journal, y = authorsCount)) +
  geom_boxplot()
```

Effect of time on number of citations.

```{r}
ggplot(counts_norm, aes(x = daysSincePublished, y = wosCountThru2011)) +
  geom_point()
```

How do the experts do?

```{r}
ggplot(counts_norm, aes(x = f1000Factor, y = wosCountThru2011)) +
  geom_point()
```

Compared to the masses?

```{r}
ggplot(counts_norm, aes(x = backtweetsCount, y = wosCountThru2011)) +
  geom_point()
```

split by journal

```{r}
ggplot(counts_norm, aes(x = backtweetsCount, y = wosCountThru2011)) +
  geom_point() +
  facet_wrap(~journal)
```

split by year

```{r}
ggplot(counts_norm, aes(x = backtweetsCount, y = wosCountThru2011)) +
  geom_point() +
  facet_wrap(~year)
```

Wikipedia and citation count

```{r}
ggplot(counts_norm, aes(x = wikipediaCites, y = wosCountThru2011)) +
  geom_point()
```

Wikipedia and Facebook

```{r}
ggplot(counts_norm, aes(x = wikipediaCites, y = facebookLikeCount)) +
  geom_point()
```

Downloads and citation counts.

```{r}
ggplot(counts_norm, aes(x = pdfDownloadsCount, y = wosCountThru2011)) +
  geom_point()
```

Downloads and mendeleyReadersCount

```{r}
ggplot(counts_norm, aes(x = pdfDownloadsCount, y = mendeleyReadersCount)) +
  geom_point()
```

HTML page view versus PDF downloads

```{r}
ggplot(counts_norm, aes(x = htmlDownloadsCount, y = pdfDownloadsCount)) +
  geom_point() +
  facet_wrap(~journal)
```

```{r}
counts_norm <- mutate(counts_norm, immuno = grepl("Immunology", plosSubjectTags))
```

```{r}
ggplot(counts_norm, aes(x = htmlDownloadsCount, y = pdfDownloadsCount,
                        color = immuno)) +
  geom_point() +
  facet_wrap(~journal)
```

Where are immuno papers published?

```{r}
ggplot(counts_norm, aes(x = journal, fill = immuno)) +
  geom_bar()
```

Evolutionary biology papers?

```{r}
counts_norm <- mutate(counts_norm, evo = grepl("Evolutionary Biology", plosSubjectTags))
```

```{r}
ggplot(counts_norm, aes(x = journal, fill = evo)) +
  geom_bar()
```
