---
layout: page
title: Intermediate programming with R
subtitle: Conditional statements
minutes: 30
---

```{r, include = FALSE}
source("tools/chunk-options.R")
opts_chunk$set(fig.path = "fig/05-conditionals-")
```

> ## Learning Objectives {.objectives}
>
> *  Filter using logical vectors created with conditional statements
> *  Search for patterns with `grepl`
> *  Make decisions with `if` and `else` statements

```{r include=FALSE}
counts_raw <- read.delim("data/counts-raw.txt.gz")
```

In the previous lesson, we were introduced to logical vectors with the functions `is.na` and `anyNA`.

```{r}
counts_raw$authorsCount[1:10]
is.na(counts_raw$authorsCount[1:10])
anyNA(counts_raw$authorsCount[1:10])
```

In this lesson we will learn how these types of logical vectors can be used for filtering data and making decisions.

### Filtering with logical vectors

Instead of providing the numbers of the rows we want, we can filter with a logical vector.

```{r}
counts_raw$authorsCount[1:10]
counts_raw$authorsCount[1:10] > 7
dim(counts_raw[counts_raw$authorsCount > 7, ])
```

Here we filtered the data to only include the `r dim(counts_raw[counts_raw$authorsCount > 7, ])[1]` rows where the number of authors was greater than 7. 

To filter for equality or non-equality, use `==` or `!=`:

```{r}
# All the articles published in the journal PLOS One
dim(counts_raw[counts_raw$journal == "pone", ])
# All the articles NOT published in the journal PLOS One
dim(counts_raw[counts_raw$journal != "pone", ])
```

Here are the other possibilities:

* `>` - "greater than"
* `<` - "less than"
* `>=` - "greater than or equal to"
* `<=` - "less than or equal to"
* `==` - "equal to"
* `!=` - "not equal to"

These logical conditions can be combined into more complex filters using the ampersand `&` ("and") or vertical bar `|` ("or") operators.

```{r}
# All the articles published in the journal PLOS One AND with more than 7 authors
dim(counts_raw[counts_raw$journal == "pone" & counts_raw$authorsCount > 7, ])
# All the articles published in the journal PLOS One OR the journal PLOS Biology
dim(counts_raw[counts_raw$journal == "pone" | counts_raw$journal == "pbio", ])
```

When we are checking one vector for multiple possibilities, it is more convenient to use the operator `%in%` instead of creating multiple "or" conditions.

```{r}
# All the articles published in the journals PLOS One, PLOS Biology, or PLOS Genetics
dim(counts_raw[counts_raw$journal %in% c("pone", "pbio", "pgen"), ])
```

Lastly, to reverse any logical vector, we can append the exclamation point `!` for "NOT".

```{r}
# All the articles NOT published in the journals PLOS One, PLOS Biology, or PLOS Genetics
dim(counts_raw[!(counts_raw$journal %in% c("pone", "pbio", "pgen")), ])
```

### Finding patterns with `grepl`

We saw in the Unix shell that we could search for lines in a file that contain a specific pattern using `grep`.
R provides similar functionality.
`grepl` searches each element of a vector for a given pattern and returns `TRUE` if it finds it, and `FALSE` otherwise.
Let's try it out using the column `plosSubjectTags`, which describes the scientific discipline(s) of the article.

```{r}
head(counts_raw$plosSubjectTags)
```

How many of the articles have to do with "Immunology"?

```{r}
dim(counts_raw[grepl("Immunology", counts_raw$plosSubjectTags), ])
```

The first argument `grepl` was the string we were searching for, and the second argument was the vector to be searched.

How many of the immunology articles were published in PLOS Medicine.

```{r}
dim(counts_raw[grepl("Immunology", counts_raw$plosSubjectTags) &
                 counts_raw$journal == "pmed", ])
```

> ## `grepl` vs. `grep` {.callout}
>
> `grepl` returns a logical vector.
> Another option is to use `grep`.
> Instead it returns the indices of the elements that contain the pattern.
> In most cases the result will be the same, but you'll have to use the correct one if you find a situation that requires only logical vectors or index positions.

### Making decisions

In addition to filtering, we can use conditional statements to adapt the behavior of the code based on the input data.
We do this using `if` and `else` statements.
The basic structure is the following:

```{r eval=FALSE}
if (condition is TRUE) {
  do something
} else {
  do a different thing
}
```

For example, we can check whether a vector contains any missing values.

```{r}
x <- counts_raw$authorsCount
if (anyNA(x)) {
  print("Be careful! The data contains missing values.")
} else {
  print("Looks good. The data does NOT contain missing values.")
}
```

Or we can check if an object is a specific data type, and convert it to the one we need.
Here we convert the column `title` from a factor to a character vector.

```{r}
x <- counts_raw$title
if (!is.character(x)) {
  x <- as.character(x)
}
```

### Challenges

> ## Filtering articles {.challenge}
>
> How many articles with the subject tag (`plosSubjectTags`) "Evolutionary Biology" were published in either PLOS One, PLOS Biology, or PLOS Medicine?

```{r include=FALSE}
dim(counts_raw[grepl("Evolutionary Biology", counts_raw$plosSubjectTags) &
                 counts_raw$journal %in% c("pone", "pbio", "pmed"), ])
```
