---
layout: page
title: Intermediate programming with R
subtitle: Knitr Chunk Options
minutes: 10
---

> ## Learning Objectives {.objectives}
> Learn how format chunks in R Markdown to display only the information you want to display. 

You've learned the basics of how to incorporate markdown syntax with code chunks in an R Markdown file. 
Let's explore some additional options for making your code chunks appear the way you want them to appear in your reports.
There are many ways to customize your chunks and you can explore all of the options by examining the [documentation](http://yihui.name/knitr/options/#chunk_options).
Here, we'll introduce you to some of the most useful options that you might use frequently.

The first thing you may want to consider is naming your code chunks, which makes degubbing easier, especially if you have a long script.
Chunk names must be unique to each chunk.

Write the name of your chunk after the {r}, like: `{r chunk_name}`

R Markdown:

<pre><code>```{r chunky_monkey}
summary(cars)
```</code></pre>

Rendered:
```{r chunky_monkey}
summary(cars)
```

You can use RStudio to navigate to chunks based on their names, which can be especially useful as your script gets long.
Click on the bottom left bar where it says `(Top Level)` and you'll see all of the chunk names in your script appear.
Additionally, naming your chunks will be beneficial to identify errors in your code or slow sections when knitting your report. 


Sometimes you may not want to see the code that produced a particular result in your report.
You can have codeblocks in your R Markdown that are evaluated, but the code is not displayed in the final report by including echo=FALSE after the `{r chunk_name}`.

R Markdown:
<pre><code>```{r funky_monkey, echo=FALSE}
summary(cars)
```</code></pre>

Rendered:
```{r funky_monkey, echo=FALSE}
summary(cars)
```

Conversely, sometimes you may want to see the code, but not the output once the code is evalutated. 
To do so, you can include `results="hide"` after the chunk_name:

R Markdown:
<pre><code>```{r brass_monkey, results="hide"}
summary(cars)
```</code></pre>

Rendered:
```{r brass_monkey, results="hide"}
summary(cars)
```

Sometimes, you may want to write a report where both the code and the output are suppressed. 
Why would you want to do that?
Perhaps you're sending a report to a collaborator and you only want them to see the final figures, but not any data manipulation steps in the middle. 
To include a chunk that is evaluated, but no output is displayed - neither the code nor the results - put `include=FALSE` after the chunk_name:

R Markdown:
<pre><code>```{r brass_section, include=FALSE}
summary(cars)
```</code></pre>

Rendered (I swear there's a chunk after this! It's just invisible!):
```{r brass_section, include=FALSE}
summary(cars)
```

There are tons of other options you can include in your chunks: sizing your figures and whether or not to display error or warning messages. 
As you write reports of your own data analysis, you can look up these options to create a report formatted in the way you want.

In addition to code chunks, you may want to include the results of an evaluation in line with regular text. For instance, you may want to describe the data in a paragraph, and include the number of individuals in that paragraph. To do so, you can indicate that a code box should be evaluated as R by including a lowercase r, such as `` `r '\x60r evaluate_me\x60'` ``

R Markdown:
```
The _cars_ dataset included in this analysis contains records for `r '\x60r dim(cars)[1]\x60'` cars. 
```

Rendered:

The _cars_ dataset included in this analysis contains records for `r dim(cars)[1]` cars. 


