---
layout: page
title: Intermediate programming with R
subtitle: Mapping data to plot aesthetics
minutes: 30
---

```{r, include = FALSE}
source("tools/chunk-options.R")
opts_chunk$set(fig.path = "fig/15-ggplot2-aes-")
```

> ## Learning Objectives {.objectives}
>
> * Map columns of data frame to plot aesthetics using `aes`
> * Add multiple layers to a plot
> * An aesthetic should be defined within `aes` only when it is data-dependent
> * Layers inherit the aesthetics defined in the original `ggplot` function call

Plotting with ggplot2 is quite different from the base R plotting functions.
While there is a decent learning curve, once you become comfortable with the basics you can easily make complicated visualizations.

```{r echo=FALSE}
counts_raw <- read.delim("data/counts-raw.txt.gz")
```

```{r echo=FALSE, message=FALSE}
library("dplyr")
```

```{r echo=FALSE}
research <- filter(counts_raw, articleType == "Research Article")
```

### A first plot

We start by loading the package.

```{r}
library("ggplot2")
```

In ggplot2, we map columns of a data frame to features of a plot, which are called aesthetics.
The base function is `ggplot`, which we supply the data frame and also the mappings.
We map the number of PDF downloads to the x variable and the number of 2011 citations to the y variable.

```{r}
p <- ggplot(data = research, mapping = aes(x = pdfDownloadsCount, y = wosCountThru2011))
```

We plot by executing the plot object.

```{r}
p
```

However we get an error stating there are no layers in the plot.
In addition to mapping aesthetics, this is the other big concept in ggplot2.
We can build a plot up in layers.
To make a simple scatter plot, we add the layer `geom_point` to the base we had created.

```{r}
p + geom_point()
```

Using the plus sign `+` we can combine multiple ggplot2 commands.
We did not need to pass any arguments to `geom_point` because it inherited the aesthetic mappings we set in the original `ggplot` command.

One option is to build up the plot in pieces by reassigning the result to `p`.

```{r}
p <- p + geom_point()
```

But so that it is clear what commands are creating each plot, we'll assign `p` in one long command.
To aid readability, each new command will go on a new line.
Also we'll remove the names of the arguments to `ggplot2` so that it is shorter.

```{r}
p <- ggplot(research, aes(x = pdfDownloadsCount, y = wosCountThru2011)) +
  geom_point()
p
```

### Exploring the different aesthetics

So far this isn't that interesting because the aesthetics `x` and `y` are so straightforward.
Let's map the column `journal` to some of the different aesthetics to see how powerful this is.
We'll define these new aesthetics within the `geom_point` layer.
We'll explore the significance of this later.

```{r}
p <- ggplot(research, aes(x = pdfDownloadsCount, y = wosCountThru2011)) +
  geom_point(aes(color = journal))
p
```

Each point was colored according the journal the article was published in.
Furthermore, a legend was automatically added to the plot.

Another aesthetic is the size of the points.
Let's assign the size of the point to the number of authors for the article.

```{r}
p <- ggplot(research, aes(x = pdfDownloadsCount, y = wosCountThru2011)) +
  geom_point(aes(size = authorsCount))
p
```

The aesthetic alpha controls the transparency of the points.
Let's map alpha to the number of days since publication to emphasize articles that have had more time to accrue citations.

```{r}
p <- ggplot(research, aes(x = pdfDownloadsCount, y = wosCountThru2011)) +
  geom_point(aes(alpha = daysSincePublished))
p
```

Up to this point it may have seemed superfluous to have the `aes` function call within the other functions.
However, this is critical for ggplot2 to know we are mapping the aesthetic to a column of the data frame.
If the aesthetic is defined outside of `aes`, then it is applied uniformly to all the data.
For example, to make all the points the color red we assign the aesthetic without `aes`.

```{r}
p <- ggplot(research, aes(x = pdfDownloadsCount, y = wosCountThru2011)) +
  geom_point(color = "red")
p
```

### Adding another layer

The plots we've made have only had one layer.
Let's return to coloring the points by the journal and add a new layer, `geom_smooth`.

```{r}
p <- ggplot(research, aes(x = pdfDownloadsCount, y = wosCountThru2011)) +
  geom_point(aes(color = journal)) +
  geom_smooth()
p
```

`geom_smooth` fits a loess curve to the data along with the 95% confidence interval.

Now let's change the aesthetics such that mapping the column `journal` to the color of the points is done in the base `ggplot` function call.

```{r}
p <- ggplot(research, aes(x = pdfDownloadsCount, y = wosCountThru2011,
                          color = journal)) +
  geom_point() +
  geom_smooth()
p
```

There are now 7 loess curves, one per journal.
Because the aesthetic color was defined in the base `ggplot` call, `geom_smooth` inherited this aesthetic mapping and correspondingly split into 7 groups.

### Challenge

> ## Citations versus days since publication {.challenge}
>
> Create a scatter plot with `daysSincePublished` mapped to the x-axis and `wosCountThru2011` mapped to the y-axis.
> Include a loess fit of the data.
> Set the transparency level (alpha) of the points to 0.5 and color the points according to the journal where the article was published.
> Make the loess curve red.

```{r include=FALSE}
p <- ggplot(research, aes(x = daysSincePublished, y = wosCountThru2011)) +
  geom_point(aes(color = journal), alpha = 0.5) +
  geom_smooth(color = "red")
p
```
