---
title: Creating subplots with facets
output: word_document
---

```{r, include = FALSE}
source("tools/chunk-options.R")
opts_chunk$set(fig.path = "fig/17-ggplot2-facets-")
```

> ## Learning Objectives {.objectives}
> *  Create subplots for each level of a factor using `facet_wrap`
> *  Create subplots for each level of a factor using `facet_grid`

In our plot of the 2011 citations versus number of PDF downloads, we colored the points according to the journal where the article was published.
However this was not very informative because the the points overlapped substantially.
In order to compare across journals, it would be better to create a separate subplot for each journal.

```{r echo=FALSE}
counts_raw <- read.delim("data/counts-raw.txt.gz")
```

```{r echo=FALSE, message=FALSE}
library("dplyr")
```

```{r echo=FALSE}
research <- filter(counts_raw, articleType == "Research Article")
```

```{r echo=FALSE, message=FALSE}
library("ggplot2")
```

```{r}
p <- ggplot(research, aes(x = log10(pdfDownloadsCount + 1),
                          y = log10(wosCountThru2011 + 1))) +
  geom_point(aes(color = journal)) +
  geom_smooth() +
  scale_x_continuous(breaks = c(1, 3), labels = c(10, 1000)) +
  scale_y_continuous(breaks = c(1, 3), labels = c(10, 1000))
p
```

### facet_wrap

We can accomplish this with the function `facet_wrap`.
We specify the name of the column after a tilde `~`, and a separate subplot is created for each level of the factor.

```{r}
p + facet_wrap(~journal)
```

Furthermore, we can control the display of the output by specifying the number of rows and/or columns.

```{r}
p + facet_wrap(~journal, ncol = 2)
```

Also notice that since there are 7 different plots, there are now seven different loess curves.

### facet_grid

If we have two categorical variables to facet, we use the function `facet_grid`.
First let's create a new column that is a logical vector for whether or not the article has the PLOS subject tag "Immunology".

```{r}
research <- mutate(research, immuno = grepl("Immunology", plosSubjectTags))
```

With `facet_grid`, we first specify the factor for the rows and second the factor for the columns.

```{r}
#p + facet_grid(journal~immuno)
```

This failed because we updated the data frame `research` with the new column `immuno`, but we never updated `p` with the new data.
Let's redefine `p` and then add use `facet_grid`.

```{r}
p <- ggplot(research, aes(x = log10(pdfDownloadsCount + 1), y = log10(wosCountThru2011 + 1))) +
  geom_point(aes(color = journal)) +
  geom_smooth() +
  scale_x_continuous(breaks = c(1, 3), labels = c(10, 1000)) +
  scale_y_continuous(breaks = c(1, 3), labels = c(10, 1000))
p + facet_grid(journal~immuno)
```

> ## Updating data {.callout}
>
> Because updating the data frame used to make a plot is so common, ggplot2 provides a shortcut.
> Instead of redefining the function, we could have updated the data frame using the operator `%+%`.
> In this case, the command would have been `p %+% research + facet_grid(journal~immuno)`.

### Challenge

> ## Using facets {.challenge}
>
> Add another variable to `research` called `evolution`, which is a logical vector indicating if the article has the PLOS subject tag "Evolutionary Biology".
> Use `facet_grid` to create subplots based on the variables `evolution` and `immuno`.

```{r include=TRUE}
research <- mutate(research,
                   evolution = grepl("Evolutionary Biology",
                                     plosSubjectTags))
p <- ggplot(research, aes(x = log10(pdfDownloadsCount + 1),
                          y = log10(wosCountThru2011 + 1))) +
  geom_point(aes(color = journal)) +
  geom_smooth() +
  scale_x_continuous(breaks = c(1, 3), labels = c(10, 1000)) +
  scale_y_continuous(breaks = c(1, 3), labels = c(10, 1000)) +
  facet_grid(evolution~immuno)
p
```
