---
layout: page
title: Intermediate programming with R
subtitle: Defensive programming with stopifnot
minutes: 30
---

```{r, include = FALSE}
source("tools/chunk-options.R")
```

> ## Learning Objectives {.objectives}
>
> * Check conditions with `stopifnot`
> * Send warnings with `warning`

As we saw in the debugging section, working functions can still produce unexpected errors.
And tracking down the cause of the errors can be difficult even with a debugging tool.
This is why the concept of defensive programming is important.
Defensive programming encourages us to frequently check conditions and throw an error if something is wrong.
These checks are referred to as assertion statements because we want to assert some condition is `TRUE` before proceeding.
This makes it easier to debug because we have a better idea of where the error originated.
Recall the uninformative error we received from `sum_calc_stat`: "Error in apply(df_sub, 1, mean): dim(X) must have a positive length", and the problem was actually introduced at the step before during the subsetting.
While it will involve more work when initially writing the code, it will save time in the future when others (including your future self) use the code.

```{r echo = FALSE}
counts_raw <- read.delim("data/counts-raw.txt.gz")
```

### Checking conditions with `stopifnot`

To create an error, we can use the function `stop`.
For example, if a variable `x` needed to be a character vector, we could check for this condition with an `if` statement and throw an error if the condition was violated.

```{r}
x <- 1
if (!is.character(x)) {
  stop("x must be a character vector.")
}
```

If we had multiple conditions to check, it would take many lines of code to check all of them.
Luckily R provides the convenience function `stopifnot`.
We can list as many requirements that should evaluate to `TRUE`, and `stopifnot` throws an error if it finds one that is `FALSE`.

```{r}
stopifnot(is.character(x))
```

If `x` also had to have a length of one, we can add that condition.

```{r}
stopifnot(length(x) == 1, is.character(x))
```

Listing these conditions also serves a secondary purpose as documentation for the code.

Let's try out defensive programming by adding assertions to check the input to our function `mean_metric_per_var`.

```{r}
mean_metric_per_var <- function(metric, variable) {
  if (!is.factor(variable)) {
    variable <- as.factor(variable)
  }
  variable <- droplevels(variable)
  result <- numeric(length = length(levels(variable)))
  names(result) <- levels(variable)
  for (v in levels(variable)) {
    result[v] <- mean(metric[variable == v], na.rm = TRUE)
  }
  return(result)
}
```

We want to assert the following:

*  `metric` is a numeric vector
*  `metric` and `variable` have the same length

```{r}
mean_metric_per_var <- function(metric, variable) {
  stopifnot(is.numeric(metric),
            length(metric) == length(variable))
  if (!is.factor(variable)) {
    variable <- as.factor(variable)
  }
  variable <- droplevels(variable)
  result <- numeric(length = length(levels(variable)))
  names(result) <- levels(variable)
  for (v in levels(variable)) {
    result[v] <- mean(metric[variable == v], na.rm = TRUE)
  }
  return(result)
}
```

It still works when given proper input.

```{r}
mean_metric_per_var(counts_raw$backtweetsCount, counts_raw$year)
```

But fails instantly if given improper input.

```{r}
# Metric is a factor instead of numeric
mean_metric_per_var(counts_raw$journal, counts_raw$year)
# The variable vector is shorter than metric
mean_metric_per_var(counts_raw$backtweetsCount, counts_raw$year[1:20])
```

We can also check the output.
Let's add an assertion to ensure that the result does not contain any `NA`s.

```{r}
mean_metric_per_var <- function(metric, variable) {
  stopifnot(is.numeric(metric),
            length(metric) == length(variable))
  if (!is.factor(variable)) {
    variable <- as.factor(variable)
  }
  variable <- droplevels(variable)
  result <- numeric(length = length(levels(variable)))
  names(result) <- levels(variable)
  for (v in levels(variable)) {
    result[v] <- mean(metric[variable == v], na.rm = TRUE)
    stopifnot(!is.na(result[v]))
  }
  return(result)
}
```

### Send warnings

In general, if something is wrong, we want to produce an error as quickly as possible to make it easier to debug.
However there are situations where it is appropriate to issue a warning to the user without stopping the code.
For example, when a non-factor is passed to `variable` in `mean_metric_per_var`, we automatically convert it to a factor.
Since the user could have simply passed in the wrong vector, we should issue a warning.
This alerts the user to double check their input.
To do this we use the function `warning`.

```{r}
mean_metric_per_var <- function(metric, variable) {
  stopifnot(is.numeric(metric),
            length(metric) == length(variable))
  if (!is.factor(variable)) {
    warning("variable was automatically converted to a factor.")
    variable <- as.factor(variable)
  }
  variable <- droplevels(variable)
  result <- numeric(length = length(levels(variable)))
  names(result) <- levels(variable)
  for (v in levels(variable)) {
    result[v] <- mean(metric[variable == v], na.rm = TRUE)
    stopifnot(!is.na(result[v]))
  }
  return(result)
}
```

Now when we use a non-factor, the function warns us that it was converted to a factor.

```{r}
mean_metric_per_var(counts_raw$backtweetsCount, counts_raw$year)
```

### Challenge

> ## Practice defensive programming {.challenge}
>
> Use defensive programming techniques to make the function `calc_sum_stat` more robust.
> 
>
> ```{r}
> calc_sum_stat <- function(df, cols) {
>   df_sub <- df[, cols, drop = FALSE]
>   sum_stat <- apply(df_sub, 1, mean)
>   return(sum_stat)
> }
> ```
>
> Some specific ideas:
>
> * Assert that the input `df` is not empty (hint: use `dim`)
> * Assert that `cols` is a character vector
> * Assert that the columns listed in `cols` are in `df` (hint: use `%in%` and `colnames`)
> * Assert that `df_sub` is a data frame (hint: use `is.data.frame`)
> * Assert that `sum_stat` is not `NA`
> * Issue a warning if `cols` only contains one column (since taking the mean of one column isn't very useful, the user may have made a mistake)
>

```{r include=FALSE}
calc_sum_stat <- function(df, cols) {
  stopifnot(dim(df) > 0,
            is.character(cols),
            cols %in% colnames(df))
  if (length(cols) == 1) {
    warning("Only one column specified. Calculating the mean will not change anything.")
  }
  df_sub <- df[, cols, drop = FALSE]
  stopifnot(is.data.frame(df_sub))
  sum_stat <- apply(df_sub, 1, mean)
  stopifnot(!is.na(sum_stat))
  return(sum_stat)
}
# Proper
sum_stat <- calc_sum_stat(counts_raw, c("wosCountThru2010", "f1000Factor"))
```

```{r include=FALSE, eval=FALSE}
# Annoyingly, errors break the build if they are in chunks where `include=FALSE`.
# Even explicitly setting `error=TRUE` does not override this behavior.

# Empty data frame
sum_stat <- calc_sum_stat(data.frame(), c("wosCountThru2010", "f1000Factor"))
# Non-character cols
sum_stat <- calc_sum_stat(counts_raw, 1:3)
# Bad column names
sum_stat <- calc_sum_stat(counts_raw, c("a", "b"))
# Issue warning since only one column
sum_stat <- calc_sum_stat(counts_raw, "mendeleyReadersCount")
# NA output
sum_stat <- calc_sum_stat(counts_raw, c("wosCountThru2010", "facebookLikeCount"))
# Should fix to `na.rm == TRUE` in call to mean
```

